Death by Dropdown?<br>Engineer Insightful Shiny Apps<br>with Behavioral Science

### 2025-09-11 conference-materials/2025_posit-conf/draft.qmd
Refactor slides for clarity, visuals, and storytelling
- Replaced `.r-fit-text` with `.text-xl` for improved readability
- Adjusted GIF embed aspect ratio for better scaling
- Polished slide text and speaker notes to be more conversational
- Added width styling to columns for balanced layouts
- Simplified "Jam Study" example, added new `jam-study.png` illustration
- Updated road trip visuals with `vw-bus.jpg` to strengthen metaphor
- Reorganized `{bidux}` slide into column-based layout for clarity
- Revised closing slide to highlight `bidux` branding instead of generic imagery
- Removed global

::: {.r-fit-text}
**Ever open a dashboard with 18 filters across 6 tabs...<br>
and wonder where to even start?**
:::

@@

::: {.column width="45%" .fragment}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.77778" data-width="100%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
::::

::: {.notes}
- 10 seconds
- Ever open a dashboard with 18 filters across 6 separate tabs...and try to figure out where to even start?
- I mean, that's a lot of data and different ways slicing it to get a full understanding of what's going on...so what do you do?
- That is...if you haven't made your career out of exploring data sets like most executives making decisions...what do you do?
- [CLICK] Maybe you do something like this monkey....just start clicking randomly hoping something sticks
- [PAUSE]
- [CLICK] Or maybe your feelings are a bit more intense...
- I get the feeling I'm not the only one who's been here before
:::

@@

:::: {.columns}
::: {.column}
### Raise hand if you've ever:
- Designed a UI with 10+ filters?
- Had a user ask for "one more dropdown"?
- Built a Shiny app with more `selectInput()` than insights?

::: {.fragment}
### Welcome! You're among friends

@@

::: {.column .fragment}
### 👋 Hi, I'm Jeremy!

@@

::: {.fragment}
- **PhD in Applied Psychology**: How people actually process information
- **Led Shiny teams at scale**: *Built dashboards people use*
- **Created {bidux} package**: *Your behavioral scientist in the console*
:::
:::
::::

@@

::: {.r-fit-text}
"Most dashboard failures aren't technical. They're psychological."
:::

@@

::: {.text-large}
The interdisciplinary study of how people **actually** make decisions<br>
(Spoiler: irrationally, predictably, consistently)
:::

:::: {.columns .fragment}
### Classic Example: The Paradox of Choice

::: {.column width="50%"}
![Jam Study Results](img/jam-study-behavioral.png){width="100%" style="border-radius: 8px; box-shadow: 0 4px 12px rgba(0,0,0,0.15);"}
:::

::: {.column width="50%"}
### Sheena Iyengar's Jam Study

::: {.text-large}
**24 varieties displayed**: 60% stopped to look<br>
**6 varieties displayed**: 40% stopped to look

But...

**24 varieties**: Only 3% bought jam<br>
**6 varieties**: 30% bought jam
:::

@@

::: {.notes}
- 1 minute
- Behavioral science studies how people think, feel, and act...often in the context of decision making
- "Irrationally but predictably"
- "Classic behavioral science example that perfectly mirrors our dashboard problem"
- "Sheena Iyengar's famous jam study at Columbia University"
- Show image: Side-by-side grocery store displays - one with 24 jam varieties (crowded, overwhelming), one with 6 varieties (clean, organized). Include simple bar charts showing the dramatic difference in purchase rates.
- "More choice attracted browsers, but paralyzed buyers"
- [CLICK] "This is exactly what happens with dashboard filters - more options kill engagement"
- "This principle drives everything we'll build with the BID framework"
:::

@@

::: {.notes}
- 20 seconds
- "Applied to dashboards"
- "These aren't just theories, they're tools"
:::

## The Solution: Your Dashboard Journey 🚗 {background-color="#2c3e50"}

:::: {.columns}
::: {.column width="30%"}
![Road Trip Adventure](img/roadtrip-journey.png){width="100%" style="border-radius: 8px; box-shadow: 0 4px 12px rgba(0,0,0,0.15);"}
:::

@@

📍 **Interpret** where you're going<br>
⚠️ **Notice** the warning lights<br>
🌦️ **Anticipate** weather changes<br>
🗺️ **Structure** the best route<br>
✅ **Validate** safe arrival

::: {.fragment}
![VW Bus Spirit](img/vw-bus-adventure.png){width="80%" style="margin-top: 20px; border-radius: 8px;"}
:::
:::

@@

![BID Hex](img/bid-framework.png){width="80%" style="margin-top: 20px;"}
:::
::::

::: {.notes}
- 45 seconds
- Show image: Scenic road stretching toward mountains with modern car, representing the journey toward better UX. Should evoke adventure, possibility, and structured progress.
- "Think of building dashboards like you would a road trip"
- "You need to interpret where you're going, notice warning lights..."
- [CLICK] Show image: Classic VW bus with hex-style stickers, embodying the "long strange trip" philosophy of exploration and community. Connects to Grateful Dead culture of innovation and shared journey.
- [CLICK] "This is the BID framework"
- "Five *iterative* stages to prevent dashboard disasters"
- "Like any good road trip, it's not just about the destination - it's about having the right framework to explore and discover along the way"
:::

@@

### Your behavioral scientist in the console

::: {.text-large}
An R package that:

@@

::: {.notes}
- 30 seconds
- "bidux is your behavioral scientist in the console"
- "Works with any Shiny app, telemetry optional, auto-suggests improvements"
- [CLICK] "Think of it as your dashboard's therapist"
:::

@@

::: {.notes}
- 1 minute
- "Notice finds the concepts behind the friction points"
- Why this stage matters for developers: It turns user complaints into actionable engineering insights
- Look at this heatmap - 14 filters completely ignored. This isn't just usage data, it's technical debt visualization
- What this means for your codebase:
    1. Those unused filters are maintenance overhead with zero ROI
    2. Database queries are running for features nobody touches
    3. UI complexity is slowing render times without value
    4. Future feature requests will compound these problems
- `bid_notice` auto-suggests behavioral theories to explain patterns, or power users can override
- This transforms "users don't like it" into "choice overload is killing task completion" - now you have a specific problem to solve with code
:::

@@

## One Thing to Remember {.center background-color="#2c3e50"}

:::: {.columns .center}

::: {.column width="30%"}
![Community Spirit](img/posit-conf-community.png){width="100%" style="border-radius: 8px; opacity: 0.8;"}
:::

@@

::: {.fragment .text-xl}
Let's fix that together!
:::
:::

::: {.column width="30%"}
![BID Hex](img/hex-bidux.png){class="hex-logo" width="200"}

::: {.fragment .text-small style="margin-top: 20px; color: var(--warm-white);"}
*Special thanks to the posit::conf organizers,<br>
the Shiny community,<br>
and all of you for caring about your users*
:::
:::
::::

>>>>

::: {.text-xl}
**Ever open a dashboard with 18 filters across 6 tabs...<br>
and wonder where to even start?**
:::

@@

::: {.column width="45%" .fragment}
<div class="tenor-gif-embed" data-postid="17038619" data-share-method="host" data-aspect-ratio="1.32447" data-width="100%"><a href="https://tenor.com/view/ron-swanson-parks-and-recreation-parks-and-rec-computer-dumpster-gif-17038619">Ron Swanson Parks And Recreation GIF</a></div> <script type="text/javascript" async src="https://tenor.com/embed.js"></script>
:::
::::

::: {.notes}
- 10 seconds
- Ever open a dashboard with 18 filters across 6 separate tabs...and try to figure out where to even start?
- I mean, that's a lot of data and different ways slicing it to get a full understanding of what's going on...so what do you do?
- That is...if you haven't made your career out of exploring data sets like most executives making decisions...what do you do?
- [CLICK] Maybe you do something like this monkey....just start clicking randomly hoping something sticks
- [PAUSE]
- [CLICK] Or maybe your feelings are a a little bigger...
- I get the feeling I'm not the only one who's been here before
:::

@@

:::: {.columns}
::: {.column width="45%"}
### Raise hand if you've ever:
- Designed a UI with 10+ filters?
- Had a user ask for "one more dropdown"?
- Built a Shiny app with more `selectInput()` than insights?

::: {.fragment}
### Welcome! You're among friends 😊

@@

::: {.column width="45%" .fragment}
### 👋 Hi, I'm Jeremy!

@@

::: {.fragment}
**PhD in Applied Psychology** → **Led Shiny teams at scale** → **Created {bidux} package**
:::
:::
::::

@@

::: {style="text-align: center; margin-top: 200px;"}
"Most dashboard failures aren't technical. They're psychological."
:::

@@

::: {.text-large}
The interdisciplinary study of how people **actually** make decisions (Spoiler: irrationally, predictably, consistently)
:::

:::: {.columns}
::: {.column width="50%"}
### Classic Example: The Paradox of Choice

::: {.text-large}
- [study details here]
:::
:::

::: {.column width="50%" .fragment}
### Study Results

![](img/jam-study.png){width="100%" style="border-radius: 8px; box-shadow: 0 4px 12px rgba(0,0,0,0.15);"}

@@

::: {.notes}
- 1 minute
- Behavioral science studies how people think, feel, and act...often in the context of decision making
- Surprise...we're not rational...but we are still predictable and fairly consistent when we pay attention to the right drivers
- Let me show you a classic example that mirrors our dashboard problem...Sheena Iyengar & Mark Lepper’s famous jam study at Columbia
- They studied people in the field at a fancy grocery store...think Whole Foods or Trader Joe's
- These shoppers walked by a display stand with either 24 jams...(extensive choice condition)...or 6 jams (limited choice condition)
- More choice drew people in...60% stopped at the big display vs. 40% at the small one
- But fewer choices drove action...nearly 30% bought jam after the 6-jam booth, compared to only 3% after the 24-jam booth
- More choice attracted browsers...but paralyzed buyers
- So what do we learn from this?
    - Too many options look appealing but kill follow-through...exactly what happens with dashboards overloaded with filters
:::

@@

::: {.notes}
- 20 seconds
- Now the paradox of choice is just one of many behavioral science concepts that directly apply to dashboards
- Taking this a step further...
- These aren't just theories...they're tools
:::

## The Solution: Your Dashboard Journey {background-color="#2c3e50"}

:::: {.columns}
::: {.column width="30%"}
![](img/vw-bus.jpg){.r-stretch}
:::

@@

📍 **Interpret** where you're going<br>
⚠️ **Notice** the warning lights<br>
🌦️ **Anticipate** weather changes<br>
🗺️ **Structure** the best route<br>
✅ **Validate** safe arrival
:::

@@

::: {.fragment}
![](img/bid-framework.png)
:::
:::
::::

::: {.notes}
- 45 seconds
- Think of building dashboards like you would a road trip
- You need to interpret where you're going, notice warning lights...
- [CLICK] "This is the BID framework"
- [CLICK] "Five *iterative* stages to prevent dashboard disasters"
- Like any good road trip, it's not just about the destination...it's about having the right framework to explore and discover along the way
:::

@@

:::: {.columns}
::: {.column width="45%" .text-large}
An R package that:

@@

::: {.column width="45%" .fragment .text-xl style="text-align: center; margin-top: 30px;"}
*Your behavioral scientist in the console*
:::
::::

::: {.notes}
- 30 seconds
- "bidux is your behavioral scientist in the console"
- "Works with any Shiny app, telemetry optional, auto-suggests improvements"
- [CLICK] "Think of it as your behavioral scientist in the console"
:::

@@

::: {.notes}
- 1 minute
- Notice finds the concepts behind the friction points
- Why this stage matters for developers...It turns user complaints into actionable engineering insights
- Look at this heatmap...14 filters basically going unused...4 are completely ignored
- This isn't just usage data...it's technical debt visualization
- What this means for your codebase:
    1. Those unused filters are maintenance overhead with zero ROI
    2. Database queries could be running for features nobody touches
    3. UI complexity is slowing rendering times without value
    4. Future feature requests will compound these problems over time
- `bid_notice` auto-suggests behavioral theories to explain patterns...or power users can override them with specific concepts
- This transforms "users don't like it" into "choice overload is killing task completion"
- Now you have a specific problem to solve with code
:::

@@

## One Thing to Remember {background-color="#2c3e50"}

:::: {.columns .center}
::: {.column width="30%"}
![](img/hex-bidux.png)
:::

@@

::: {.fragment .text-xl style="margin-top: 50px"}
Let's fix that together!
:::
:::

::: {.column width="30%" .fragment}
*Special thanks to the posit::conf organizers,<br>
the Shiny community,<br>
and all of you for caring about your users*
:::
::::
